---
title: "Day 4 Prac Statistical Computing"
author: "GLDNAT011 Natalie Gelderblom"
format: html
---

## Introduction

This practical aims to draw insight through rigourous statistics using the tidyverse and use `tidyr` and `dplyr` wherever appropriate. It will be based on the [nycflights13](https://cran.r-project.org/web/packages/nycflights13/index.html) dataset (which contains information on over 300,000 flights from New York City in 2013) and touch on a case study using baseball batting averages from the Lahman package.

To set up, run the code below:

```{r}
# Install tidyverse if not already installed 

if (!requireNamespace("tidyverse", quietly = TRUE)) {
  install.packages("tidyverse")
}
library(tidyverse)  

# Install and load nycflights13 for flight data
if (!requireNamespace("nycflights13", quietly = TRUE)) {
  install.packages("nycflights13")
}

data(flights, package = "nycflights13")
flights
```

## **Questions**

Using the `flights` dataset:

1.  Display the `flights` dataset in an alternative format to simply printing it (i.e. running `flights`).

```{r}

# two options 

head(flights)

glimpse((flights))

```

2.  Rewrite this code using `dplyr` and the pipe:

This code creates a tibble of all of the means and standard deviations of the carriers for month 1 and then orders it according to the mean.

```{r}
flight1 <- flights[flights$month == 1, ]
carrier_vec <- unique(flight1$carrier)
carrier_dist_vec_mean <- numeric(length(carrier_vec))
carrier_dist_vec_sd <- numeric(length(carrier_vec))

for (i in seq_along(carrier_vec)) {
  carrier_dist_vec_mean[i] <- mean(
    flight1$distance[flight1$carrier == carrier_vec[i]]
   )
  carrier_dist_vec_sd[i] <- sd(
    flight1$distance[flight1$carrier == carrier_vec[i]]
  )
}
dist_tbl <- tibble(
  carrier = carrier_vec,
  mean_distance = carrier_dist_vec_mean,
  sd_distance = carrier_dist_vec_sd
)
dist_tbl[order(dist_tbl$mean_distance), ]
```

This code does the same thing just using the pipe and dplyr.

```{r}

# Extract all of the flights in month 1 and group them by carrier type 
flights.1.c <- flights |> filter(month==1) |> group_by(carrier) 

# Get the mean and the sd of distance by each carrier
summary.stats <- flights.1.c |> summarise(mean_dist = mean(distance),sd_dist = sd(distance))

# sort the tibble of the carrier means

distance.bymean.sd <- summary.stats |> arrange(mean_dist)

distance.bymean.sd

```

3.  Explain why the standard deviation is `NA` for one carrier, and why it is `0` for others. Demonstrate your answer using code.

There is only 1 observation for OO carrier in month 1 so the standard deviation shows as NA.

```{r}
# ================= Why is OO sd NA? ================
flights.OO.1 <- flights |> filter(carrier=="OO") |> filter(month==1)
flights.OO.1

# Only 1 observation for OO carrier in month 1 so the standard deviation shows as NA
nrow(flights.OO.1)
```

All flights by YV planes in month 1 fly the same distance of 229 so there is no variation in the distance traveled by YV carrier flights in month 1. This results in a standard deviation of zero.

These observations are demonstrated using the code below:

```{r}
# ================= Why is YV sd zero? ================

flights.YV.1 <- flights |> filter(carrier=="YV") |> filter(month==1)
flights.YV.1
distinct.distances <- distinct(flights.YV.1,distance)
distinct.distances

# All flights by YV planes in month 1 fly the same distance of 229
# there is no variation in the distance travelled by YV carrier flights in month 1 
```

4.  Using `tidyr` and `dplyr` where appropriate, construct a dataframe where the carriers are along the columns, and the rows are the average departure delay (`dep_delay`) flown by each carrier (`carrier`) in each month.

```{r}

# find the average departure delay (dep_delay) flown by each carrier (carrier) in each month
av_dep_delay <-  flights |> group_by(carrier,month) |> summarise(average.dep.delay = mean(dep_delay, na.rm=TRUE)) 
av_dep_delay 

# construct a dataframe where the carriers are along the columns, and the rows are the average departure delay in each month
by_carrier_df <- av_dep_delay|> pivot_wider(names_from = carrier, values_from = average.dep.delay)
by_carrier_df

```

4.  Calculate the proportion of flights that were delayed (`dep_delay` greater than 0) but arrived on or before time (`arr_delay` less than or equal to 0).

    ```{r}

    # extract the flights which meet the criteria 
    flights.relevant <- flights |> filter(dep_delay > 0 & arr_delay <=0) 
    flights.relevant

    # calculate the proportion of flights 
    proportion <- nrow(flights.relevant) / nrow(flights)
    proportion

    ```

5.  Using the `airlines` and `flights` datasets, do the following, showing the output from each step:

-   Identify routes that more than one airline flies

```{r}

# filter the rows so that it is a data frame with each carrier for each route
routes.withcarriers <- flights |> group_by(origin,dest) |> distinct(carrier, .keep_all=TRUE)
routes.withcarriers

# select the routes that have multiple carriers 
multiple.carriers <- routes.withcarriers |> tally() |> filter(n>1)
multiple.carriers

```

-   For each such route, calculate the average arrival delay for each airline (exclude NAs). Find the names of these airlines.

```{r}

# extract the flights that are from the routes in the multiple carriers data frame 
# innerjoin shows all of the entries in the first table that have matches in the second table 
mulit.carriers <- flights |> inner_join(multiple.carriers)
mulit.carriers

# calculate the average arrival delay
av_ar_delay <- mulit.carriers |> group_by(origin,dest,carrier) |> summarise("average_arrival_delay"=mean(arr_delay, na.rm=TRUE)) 
av_ar_delay


# Find the names of these airlines
airlines = nycflights13::airlines

with_names <- left_join(av_ar_delay,airlines)
with_names

```

-   For each such route, identify the airline with the worst and best average arrival delay.

```{r}

# using the summarise fucntion find the minimum and maximum delay and have the relevant airline next to their average delay time 
min.max <- with_names|> group_by(origin,dest) |> summarise("min.d" =min(average_arrival_delay),"smallest_delay"=name[average_arrival_delay==min(average_arrival_delay)], "max.d" = max(average_arrival_delay), "biggest_delay"=name[average_arrival_delay==max(average_arrival_delay)]) 
min.max
```

-   Identify the route with the greatest difference between the best and worst performing airlines

```{r}
ranges <- min.max|> group_by(origin,dest) |> mutate(range.d = max.d-min.d,.before = 1) 
ranges


biggest_range <-ranges |> arrange()
origin_biggest_delay <-biggest_range$origin[1]
dest_biggest_delay <-biggest_range$dest[1]
```

The route with the greatest difference between the best and the worst performing airlines was from `origin_biggest_delay` to `dest_biggest_delay`

-   Determine the reason for this difference

```{r}
info <- flights |> filter(origin=="EWR",dest=="ATL") 
info

info.2 <- flights |> filter(origin=="EWR",dest=="ATL") |> distinct(carrier)
info.2
# there are 4 carriers  DL, EV, UA, 9E

how_many_flights <- info |> group_by(carrier)|> tally() |>arrange(n)
how_many_flights

most_flights <- how_many_flights[1,]
most_flights

least_flights <- how_many_flights[4,]
least_flights

# most flights is 9E
# lest flights is 

best_airline <- biggest_range[1,5]
best_airline #  Endeavor Air Inc. 9E 

worst_airline <-biggest_range[1,7]
worst_airline # express jet EV


summary(info)

summary(flights)
```

The best airline in the route has far fewer flights on that route so it has less delays overall. It is a very small amount of flights for the best airline (EV = Endeavor Air Inc. ) so it is not nessesarily representative. It is also interesting that the airline with the most flights (DL = Delta Air Lines Inc. ) is not the airline with the most delays.

7.  Identify all columns with missing entries, typos and any other inconsistencies in the dataset below (load it just by running the code; created using `dput` command, FYI

```{r}

# load the data 

data.2 <- structure(list(id = c("id_1", "id_2", "id_3", "id_4", "id_5", 
"id_6", "id_7", "id_8", "id_9", "id_10", "id_11", "id_12", "id_13", 
"id_14", "id_15", "id_16", "id_17", "id_18", "id_19", "id_20", 
"id_21", "id_22", "id_23", "id_24", "id_25", "id_26", "id_27", 
"id_28", "id_29", "id_30", "id_31", "id_32", "id_33", "id_34", 
"id_35", "id_36", "id_37", "id_38", "id_39", "id_40", "id_41", 
"id_42", "id_43", "id_44", "id_45", "id_46", "id_47", "id_48", 
"id_49", "id_50"), age = c(50L, 34L, 70L, 33L, 22L, 61L, 69L, 
73L, 62L, 56L, 71L, 33L, 73L, 44L, 45L, 46L, 24L, 70L, 46L, 76L, 
47L, 76L, 28L, 48L, 54L, 27L, 45L, 26L, 61L, 28L, 38L, 55L, 33L, 
36L, 62L, 58L, 72L, 31L, 34L, 51L, 61L, 64L, 26L, 28L, 60L, 29L, 
42L, 46L, 79L, 72L), gender = c("male", "male", "male", "female", 
"female", "male", "female", "male", "male", "female", "female", 
"male", "male", "female", "male", "male", "male", "male", "female", 
"male", "male", "male", "male", "female", "femal", "male", "female", 
"female", "female", "female", "male", "female", "female", "female", 
"male", "male", "female", "male", "female", "female", "male", 
"female", "female", "male", "male", "female", "male", "male", 
"male", "female"), height = c(174.4, 197.7, 174.1, 194.5, NA, 
180.4, 170.5, 157.4, 196.8, 165.1, 153, 197.4, 186, 157.1, 177.5, 
197.7, 179.3, 170.2, 182.4, NA, 165.4, 161, 168.5, 199.2, 157.7, 
154.6, 157.1, 184.5, 181, 194.6, 183.6, 186.9, 176.1, 183, 191.1, 
189.3, 199, 172, 165.6, 170.5, 150.5, 159.2, 192.1, 161.6, 162, 
153.8, 162.3, 186.6, 192.4, 174.9), weight = c(69.4, 62.3, 55.6, 
69.5, 78.6, 60.8, 72.2, 60.9, 75.1, 67.7, 82.5, 68.7, 67.8, 76.7, 
87, 61.1, 70.6, 63.3, 81.5, 59.2, 93.2, 87.3, 83.4, 80.9, 68.6, 
76.5, 93.7, 79.1, 92, 65.6, 85.4, 63.3, 79.7, 74.1, 63.3, 78.2, 
95.7, 95.1, 63.7, 66.1, 99.3, 81, 96.9, 73.3, 70.3, 83, 57.6, 
78.6, 61.9, 98.1), blood_type = c("O", "A", "O", "O", "B", "AB", 
"O", "O", "O", "AB", "A", "O", "O", "O", "B", "A", "B", "AB", 
"O", "AB", "A", "AB", "O", "B", "A", "A", "B", "AB", "A", "B", 
"B", "A", "O", "O", "O", "B", "O", "A", "A", "B", "A", "O", "AB", 
"A", "A", "O", "O", "B", "A", "O"), disease_status = c("diseased", 
"healthy", "healthy", "healthy", "healthy", "healthy", "diseased", 
"healthy", "diseased", "Healthy", "diseased", "healthy", "diseased", 
"healthy", "diseased", "healthy", "healthy", "healthy", "healthy", 
"healthy", "healthy", "diseased", "healthy", "diseased", "healthy", 
"healthy", "healthy", "healthy", "diseased", "diseased", "healthy", 
"healthy", "healthy", "diseased", "diseased", "diseased", "healthy", 
"diseased", "healthy", "healthy", "healthy", "healthy", "healthy", 
"diseased", "diseased", "diseased", "healthy", "healthy", "diseased", 
"diseased"), cholesterol = c(228, 223, 213, 198, 166, 151, 195, 
199, 189, 196, 221, 156, 185, 230, 234, 174, 185, 236, 235, 180, 
165, 220, 160, 153, 250, 153, 184, 242, 212, 179, 224, 233, 181, 
199, 220, 214, 214, 248, 191, 162, 203, 173, 199, 187, 248, 189, 
173, 212, 164, 247), glucose = c(96, 78, 101, 119, 103, 91, 86, 
NA, 77, 80, 115, 85, 88, 109, NA, 71, 90, 94, 91, 87, 113, 93, 
97, 118, 109, 80, 85, 119, 99, 108, 89, 108, 97, 116, 79, 84, 
75, 81, 119, NA, 106, 109, 75, 82, 84, 75, 76, 120, 119, 77), 
    smoker = c("yes", "yes", "yes", "yes", "no", "yes", "no", 
    "yes", "no", "no", "no", "no", "no", "yes", "no", "yes", 
    "yes", "yes", "yes", "yes", "yes", "yes", "yes", "yes", "no", 
    "no", "yes", "yes", "yes", "no", "no", "yes", "no", "yes", 
    "no", "yes", "no", "yes", "yes", "yes", "no", "no", "yes", 
    "no", "no", "no", "no", "no", "no", "yes"), exercise = c("occasional", 
    "regular", "occasional", "regular", "none", "occasional", 
    "regular", "none", "occasional", "none", "occasional", "none", 
    "none", "regular", "occasional", "none", "regular", "regular", 
    "none", "occasional", "none", "occasional", "occasional", 
    "occasional", "regular", "occasional", "regular", "regular", 
    "regular", "occasional", "occasional", "none", "none", "regular", 
    "occasional", "occasional", "none", "none", "none", "none", 
    "occasional", "regular", "regular", "none", "regular", "occasional", 
    "occasional", "none", "occasional", "regular"), income = c(84820L, 
    81547L, 22588L, 72490L, 74533L, 25338L, 41469L, 57315L, 63629L, 
    88662L, 62615L, 56261L, 58499L, 82232L, 77584L, 77275L, 38468L, 
    54510L, 91326L, 78611L, 31402L, 29586L, 21441L, 58269L, 84173L, 
    88295L, 37940L, 43750L, 69750L, 92356L, 82518L, 91455L, 68866L, 
    51178L, 68275L, 27689L, 35418L, 81318L, 62405L, 86851L, 25654L, 
    47553L, 74474L, 51409L, 22607L, 55360L, 96351L, 21516L, 41927L, 
    55810L), education = c("master", "bachelor", "PhD", "master", 
    "bachelor", "highschool", "PhD", "highschool", "PhD", "PhD", 
    "bachelor", "highschool", "master", "bachelor", "PhD", "PhD", 
    "PhD", "bachelor", "master", "highschool", "PhD", "highschool", 
    "bachelor", "master", "highschool", "highschool", "master", 
    "master", "bachelor", "PhD", "highschool", "PhD", "master", 
    "master", "master", "PhD", "highschool", "master", "master", 
    "highschool", "bachelor", "highschool", "bachelor", "PhD", 
    "bachelor", "highschool", "master", "highschool", "bachelor", 
    "bachelor"), region = c("North", "South", "North", "West", 
    "North", "West", "South", "South", "West", "South", "West", 
    "South", "West", "East", "North", "West", "North", "North", 
    "West", "North", "East", "West", "South", "North", "North", 
    "East", "East", "North", "North", "West", "South", "West", 
    "West", "East", "West", "North", "West", "North", "East", 
    "North", "West", "South", "South", "East", "North", "West", 
    "West", "East", "North", "East"), marital_status = c("divorced", 
    "single", "divorced", "divorced", "divorced", "divorced", 
    "divorced", "married", "divorced", "married", "divorced", 
    "widowed", "married", "single", "widowed", "widowed", "single", 
    "divorced", "widowed", "widowed", "single", "married", "single", 
    "married", "widowed", "married", "single", "single", "widowed", 
    "married", "widowed", "divorced", "single", "married", "single", 
    "widowed", "widowed", "married", "widowed", "divorced", "married", 
    "married", "divorced", "single", "married", "widowed", "divorced", 
    "divorced", "single", "divorced")), row.names = c(NA, -50L
), class = c("tbl_df", "tbl", "data.frame"))
```

```{r}

# Identify all columns with missing entries, typos and any other inconsistencies in the dataset

library(skimr)
skim(data.2)


# install the UtilsDataRSV package, use the following code:

if (!requireNamespace("remotes", quietly = TRUE)) {
  install.packages("remotes")
}
remotes::install_github("SATVILab/UtilsDataRSV")
library(UtilsDataRSV)
view_cols(data.2)
```

Missing entries, typos and any other inconsistencies:

1\. gender

femal is a type for female

2\. height

There are NA (missing) values in height

3\. healthy

Healthy" and "healthy should just be one thing

Would make Healthy into healthy since everything else is lower case

4.  glucose

Has NA (missing) values
