---
title: "GLDNAT011 Natalie Gelderblom Prac1 StatsComp"
---

## Question 1

Airquality is a built in R data set of daily air quality measurements in New York, May to September 1973. The first few observations are shown below:

```{r}
head(airquality)
```

1.Find all rows in “airquality” that have missing values.

```{r}

rows.NA <-which(is.na(airquality)) # get the indices of all the rows which have NA
airquality[rows.NA,] # print out the corresponding rows 

```

## Question 2

2\. Find mean, sd, min, max for each of temperature and ozone level.

```{r}

#-------- define your variables----------------------------
# Ozone has missing variables so you have to clean it before you can use the other functions like mean()
ozone.cleaned <-na.omit(airquality$Ozone) # na.omit() returns object without incomplete cases
temperature <-airquality$Temp

#------------Get mean, sd, min, max---------------

# apply(data, rows-1 cols - 2, function with no brackets)
 
# For temperature
t.mean<-mean(temperature) # temp |> mean(na.omt) 
t.sd <-sd(temperature)
t.max <-max(temperature) 
t.min <-min(temperature)

# for ozone
o.mean<-mean(ozone.cleaned) 
o.sd <-sd(ozone.cleaned)
o.max<-max(ozone.cleaned) 
o.min<-min(ozone.cleaned)

#---------------create a table to display it------------------

# create a matrix
tmp.ozn.data <-matrix(c(t.mean,t.sd,t.max,t.min,o.mean,o.sd,o.max,o.min),ncol=4,byrow=TRUE)

#specify column and row names 
colnames(tmp.ozn.data) <-c("mean","sd","min","max")
rownames(tmp.ozn.data) <-c("temperature","ozone")

# make it a table and display
as.table(tmp.ozn.data)
```


## Question 3

3\. For linear regression, parameter estimates can be found as follows:

$\hat{\beta}  = (X^TX)^{-1}X^TY$

Where Y is the response variable and X is the design matrix.

In the cars data set the response variable Y is stopping distance and the independent variable is speed.

Using the equation above you get the following results for the intercept and coefficient estimates:

```{r}

#--------------define variables---------------------------------------
X = cbind(1, cars$speed)
Y = cars$dist
# design matrix needs to be a 2 column matrix to have the output calculate both the intercept and the coefficient

#-------------Matrix Operations-----------------------------------------
# t(matrix) tranpose of a matrix 
# v1 %*% m1 matrix multiplication
# solve(m1) inverse of matrix 

#---------Find estimates using equation ----------------------------

beta_hats = solve(t(X) %*% X) %*% t(X) %*% Y

beta_intercept = beta_hats[1,1] # extract the intercept from the beta_hats matrix
beta_coefficent =beta_hats[2,1] # extract the speed coeffcient from the beta_hats matrix

print(paste("the intercept is", beta_intercept,"and the coefficent on speed is", beta_coefficent))
```

## Question 4

4.  You can check that you get the same estimates as when fitting the linear regression model using lm() in R.

```{r}
# lm(dependent~independent,data=?)
lm.model <-summary(lm(dist ~ speed, data = cars))
lm.model
```

