---
title: "Jakes Prac Draft"
author: "Natalie Gelderblom"
format: html
---

## Lowess Smoothing

Locally wieighted scatterplot smoothing (Lowess) is a smoothing technique which is a mixture of moving averages and local regression. Below, we briefly outline the technique before creating a customLowess function.

#### What is Lowess?

The Lowess approach attempts to look at the closest neighbours of a data point to find some weighted average of the closest neighbours to act as the smoothed data point, providing a smoothed function \$ \hat{y} = f(x) \$

## **Steps to Lowess**

### **1. Data**

As usual, we have observed data pairs(x1,y1),…,(xn,yn).

### **2. Span**

We need to choose the **span** of the smoothing, denoted f  such that 0\<f\<1. The span defines the number of closest neighbours to use when smoothing point xi such that

k=⌈f⋅n⌉

where k is the number of closest neighbours used for point xi and n is the number of observations.

### **3. Compute Weights**

The weights applied to each neighbour is calculated using the tricube kernel such that

wj=(1−(\|xj−xi\|dmax)3)3

$$
w_j = ( (1-(\frac{|x_j-x_i|}{d_{max}})^3 )^3
$$

where wj is the weight for neighbour xj, 

dmax is the distance to the farthest neighbour among the k neighbours selected.

Note that this means that if \|xj−wi\|≥dmax, then wj=0.

So if the distance of the observation is further than the dmax distance then it has a weight of zero and does not impact the regression.

### **4. Weighted Regression**

Perform a weighted linear regression

y=β0+β1x+e

where the weighting means that we want to minimise∑j∈neighbourswj(yj−(β0+β1xj)2)

$$
\sum_{j \in neighbours} w_j ( y_i - (\beta_0 + \beta_1x_i)^2 ) 
$$

The optimal β is given by

$$
\hat{\beta} = (X^TWX)^{-1}X^TWy
$$

where W is the diagonal matrix given by the diagonal values (w1,...,wk).

The smoothed value for xi is then the predicted value

$$
\hat{y_i} = \hat{\beta_0} + \hat{\beta_1}x_i
$$

### **5. Repeat for each** xi

## **Practical**

1.  **Generate Simulated Data:**

    -   Set your seed to 1, i.e. `set.seed(1)`.

    ```{r}
    set.seed(1)
    ```

    -   Create x as a sequence of numbers from 1 to 100.

    ```{r}
    x <- seq(1:100)
    x
    ```

    -   Generate y as a noisy sine wave such thatyi=sin⁡(xi10)+ei

        where \^ei∼N(0,0.2)\^2 )

$$
y_i = sin(\frac{x_i}{10})+ e_i
$$

```{r}

# generate 100 random errors, with a normal distribution with a mean of 0 and a standard deviation of 0.22
e <- rnorm(100,mean=0,sd=(0.2)^2)
e

# idk if it should be 100


y = sin(x/10)+ e
plot(y)


```

1.  **Implement the LOWESS Algorithm:**

    -   Define a function `customLowess(x, y, f)` that returns the smoothed values.

    0\<f\<1

```{r}
library(dplyr)
```

# ask tutors??

why does it need to only be a small matrix?

why not work with 100 x 100 ?

for W and X

```{r}

customLowess <- function(x,y,f) {
  
  #f=0.2
  # find the number of observations in your data
  n=length(x)
  
  # create a vector to hold the fitted values to return 
  yfitted<- numeric(n)
  
  weights <- cbind(1:n,numeric(n))
  weights
  
  
  
  
  
  #==========================get the span ==========================
  # k is the number of closest neighours used for xi
  # the notation in th q means ceiling round 
  k=ceiling(f*n)

  
  
  # ---------------loop through each observation------------------
  for (i in 1:n) {
    
    #i=20
  
    #====================calculate the weights =====================
    
    xi <- x[i] # extract the current observation 
    distance_vec<- abs(x-xi) # distance between observations and the current x
    
    # get the dmax 
    distances.df <- data.frame(x,distance_vec,y) |> arrange(distance_vec)
    
    dmax <- distances.df[k,2]
    
    # take the k closest observations to xi 
    k_neighbours <- distances.df[1:k,]
    k_neighbours
    
    # get the weights
    # w <- cbind(k_neighbours[,1,],(1-(k_neighbours[,2]/dmax)^3)^3)
    
    w = numeric(n)
    w[k_neighbours$x] <- (1-(k_neighbours[,2]/dmax)^3)^3
    w
    
    #k_neighbours_indices<- distances.df[1:k,1]
    #k_neighbours_indices
    
    
    # another way to do this? 
    
    #for (i in k_neighbours_indices){
    #  weights[i,2]<-w[i]
    #}
   
    
    #==================== do the weighted regression ================
    
    # make the W matrix with weights as the diagonal 
    W <- diag(w)
    W
    
    # X is the design matrix so there must be both observations and 1 
    X <- cbind(1,x)
    X

    # find the The optimal beta
    beta_hats = solve(t(X) %*% W %*% X) %*% t(X) %*% W %*% y
    beta_hats
   
    # calculate the fitted value and put it into yfitted[i]
    yfitted[i] <- beta_hats[1,1] + beta_hats[2,1]*xi
  }
  
  #=================return the predicted values====================
  return(yfitted)
}




fitted_y <- customLowess(x,y,0.2)
plot(fitted_y)


```

```{r}




# =============================working out ==================================

customLowess <- function(x,y,f) {
  
  # take this out its just for now to test stuff 
  f=2/3
  
  # create a vector to hold the fitted values to return 
  yfitted<- numeric(n)
  
  # find the number of observations in your data
  n=length(x)
  
  
  #============get the span ===================
  # k is the number of closest neighours used for xi
  k=round(f*n)
  
 
  
  #==========compute the weights=============
  
  # xi is observation or point 
  # xj is the neighbour to xi
  # weight wj = for neighbour xj

  
  # ---------------loop through each observation------------------
  for (i in 1:n) {
    
    i=4
    xi <- x[i] # extract the current observation 
    distance_vec<- abs(x-xi) # distance between observations and the current x
    distance_vec
    
    distances.df <- data.frame(x,distance_vec) |>sort()
    distances.df
    
    distances.sorted <- distances.df |> sort(distance_vec)
    distances.sorted
    
    
    
    # vectors to be filled 
    
    w <- numeric(n) # for weights 
  
  #------------find the distances to all other observations-----
    
    
    
    for (j in 1:n)
  {

    # find all the distances and sort it 
      # need to make it a vector
      dist_vec[j]<- abs(x[j]-x[i])

    } 

    
    #------------------- get dmax --------------------
    
    # find the kth distance to get dmax
    distances_sorted <- dist_vec |> sort() 
    dmax <- distances_sorted[k]

    
    
    #---------- calculate the weights -------------------
    for (j in 1:n){
      
      # if the distance is more than dmax then the weight is just zero
      if (dist_vec[j]>= dmax)
        {
          w[j]=0
  # else the weight is calculated using the formula 
    } else{
      w[j] = (  1-  ( abs(x[j]-x[i]) / dmax )^3   )^3
    }
      
      #==============================================================
      #=================correct until here!!!========================
      #++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
  

    }
    #==================== do the weighted regression ================
    # W = diagonal matrix with diagonal given by the weights
    W <- diag(w,n,n)
    print("this is w")
    print(W)
    
    # X is the design matrix so there must be both observations and 1 
    X <- cbind(1,x)
    
    print("this is x")
    print(X)
    ym <- as.matrix
    # find the The optimal beta
    beta_hats = solve(t(X) %*% W %*% X) %*% t(X) %*% W %*% ym
    beta_intercept = beta_hats[1,1] # extract the intercept from the beta_hats matrix
    beta_coefficent =beta_hats[2,1] # extract the speed coeffcient from the beta_hats matrix
    
    #======== return the predicted value=============
    yfitted[i] <- beta_intercept + beta_coefficent*x[i]

    
  } # this is the final one for each observation 
  
  return(yfitted)
 
} # this is the final one for the function
    


```

```{r}
customLowess(x) 
```

1.  **Compare with R’s Built-in `lowess()`:**

    -   Use the built-in `lowess()` function with the same `f` value. You will also need to set the `iter` argument to 0.

    -   Plot both curves to compare their smoothing values.

```{r}

plot(y)

R.lowess <- lowess(x,y,0.2,iter=0)

lines(R.lowess,col="green",lwd=10)
lines(fitted_y,col="red",lwd=2)
```
