---
title: "Prac 2"
author: "Natalie Gelderblom"
format: html
---

## Introduction

This practical aims to draw insight through rigourous statistics using the tidyverse and use `tidyr` and `dplyr` wherever appropriate. It will be based on the [nycflights13](https://cran.r-project.org/web/packages/nycflights13/index.html) dataset (which contains information on over 300,000 flights from New York City in 2013) and touch on a case study using baseball batting averages from the Lahman package.

To set up, run the code below:

```{r}
# Install tidyverse if not already installed 

if (!requireNamespace("tidyverse", quietly = TRUE)) {
  install.packages("tidyverse")
}
library(tidyverse)  

# Install and load nycflights13 for flight data
if (!requireNamespace("nycflights13", quietly = TRUE)) {
  install.packages("nycflights13")
}

data(flights, package = "nycflights13")
flights
```

## **Questions**

Using the `flights` dataset:

1.  Display the `flights` dataset in an alternative format to simply printing it (i.e. running `flights`).

2.  Rewrite this code using `dplyr` and the pipe:

```         
flight1 <- flights[flights$month == 1, ] carrier_vec <- unique(flight1$carrier) carrier_dist_vec_mean <- numeric(length(carrier_vec)) carrier_dist_vec_sd <- numeric(length(carrier_vec)) for (i in seq_along(carrier_vec)) {   carrier_dist_vec_mean[i] <- mean(     flight1$distance[flight1$carrier == carrier_vec[i]]    )   carrier_dist_vec_sd[i] <- sd(     flight1$distance[flight1$carrier == carrier_vec[i]]   ) } dist_tbl <- tibble(   carrier = carrier_vec,   mean_distance = carrier_dist_vec_mean,   sd_distance = carrier_dist_vec_sd ) dist_tbl[order(dist_tbl$mean_distance), 
```

```{r}

# ==============understanding the question and what the code is doing ===================

# extract all of the flights where the month is 1 
flight1 <- flights[flights$month == 1, ] 

# create a vector which extracts all of the unique carrier values from flight1
carrier_vec <- unique(flight1$carrier)

# numeric() generates object with a numeric class of the length of the interger put in
# of length the same as carrier_vec

carrier_dist_vec_mean <- numeric(length(carrier_vec)) # 16 zeros
carrier_dist_vec_sd <- numeric(length(carrier_vec)) # 16 zeros


# seq_along is for generating regular sequences of the same length as the object passed to it 

for (i in seq_along(carrier_vec)) {
  
  # replace the ith element of carrier_dist_vec_mean with the mean of...
  carrier_dist_vec_mean[i] <- mean(
    # the distance when carrier is equal to the ith element of the carrier vector 
    flight1$distance[flight1$carrier == carrier_vec[i]]
   )
  
  # replace the ith element in c_d_v_s with the sd of the distance 
  # when the flifht varrier is equal to the ith element of the carrier vector 
  # create an object which contains the the sd of the distances for each carrier 
  carrier_dist_vec_sd[i] <- sd(
    flight1$distance[flight1$carrier == carrier_vec[i]]
  )
}

# create a tibble of the carrier, mean distance and sd of distance 
dist_tbl <- tibble(
  carrier = carrier_vec,
  mean_distance = carrier_dist_vec_mean,
  sd_distance = carrier_dist_vec_sd
)

# order the tibble according to mean distance 
dist_tbl[order(dist_tbl$mean_distance), ]
```

```{r}


# -------------------------------working out for the question --------------




# ===========extract all of the flights where the month is 1================
# filter keeps all of the rows which meet a given criteria 
flight1.np <- filter(flights,month==1)

#using pipe 
flight1. |> fliter(month==1)




# ======create a vector which extracts all of the unique carrier types from flight1========

# made a df not a vector I hope thats fine 
carrier.categories.np <- distinct(flight1.,carrier)

# using pipe 
carrier.categories <- flight1. |> distinct(carrier)
carrier.categories





sum.stats <- flights |> group_by(carrier) |> summarise(avg_dist = mean(distance, na.rm = TRUE))
sum.stats


# numeric() generates object with a numeric class of the length of the interger put in
# of length the same as carrier_vec

carrier_dist_vec_mean <- numeric(length(carrier_vec)) # 16 zeros
carrier_dist_vec_sd <- numeric(length(carrier_vec)) # 16 zeros


# seq_along is for generating regular sequences of the same length as the object passed to it 

for (i in seq_along(carrier_vec)) {
  
  # replace the ith element of carrier_dist_vec_mean with the mean of...
  carrier_dist_vec_mean[i] <- mean(
    # the distance when carrier is equal to the ith element of the carrier vector 
    flight1$distance[flight1$carrier == carrier_vec[i]]
   )
  
  # replace the ith element in c_d_v_s with the sd of the distance 
  # when the flifht varrier is equal to the ith element of the carrier vector 
  # create an object which contains the the sd of the distances for each carrier 
  carrier_dist_vec_sd[i] <- sd(
    flight1$distance[flight1$carrier == carrier_vec[i]]
  )
}

# create a tibble of the carrier, mean distance and sd of distance 
dist_tbl <- tibble(
  carrier = carrier_vec,
  mean_distance = carrier_dist_vec_mean,
  sd_distance = carrier_dist_vec_sd
)

# order the tibble according to mean distance 
dist_tbl[order(dist_tbl$mean_distance), ]


flit.xx <-group_by(flight1,carrier)
flit.xxy <- flight1 |> group_by(carrier) |> summarise(avg_distance = mean(distance, na.rm = TRUE))
# na.rm = true means exclude 
flit.xxy

flit.xxyzx <- flights |> filter(month==1) |> group_by(carrier) |> summarise(avg_distance = mean(distance, na.rm = TRUE),sd_dist = sd(distance))
flit.xxyzx 

new.flit <- arrange(flit.xxyzx ,avg_distance)
new.flit

```

```{r}

# Extract all of the flights in month 1 and group them by carrier type 
flights.1.c <- flights |> filter(month==1) |> group_by(carrier) 

# Get the mean and the sd of distance by each carrier
summary.stats <- flights.1.c |> summarise(mean_dist = mean(distance),sd_dist = sd(distance))

# sort the tibble of the carrier means 
distance.bymean.sd <- summary.stats |> arrange(mean_dist)

distance.bymean.sd

```

3.  Explain why the standard deviation is `NA` for one carrier, and why it is `0` for others. Demonstrate your answer using code.

```{r}



# ================= Why is OO sd NA? ================

# this gives a non NA sd so idk why? 

flights.OO.1 <- flights |> filter(carrier=="OO") |> filter(month==1)
flights.OO.1

# there is only 1 observation for OO carrier in month 1 so the standard deviation shows as NA
nrow(flights.OO.1)




# ================= Why is YV sd zero? ================


flights.YV.1 <- flights |> filter(carrier=="YV") |> filter(month==1)
flights.YV.1
distinct.distances <- distinct(flights.YV.1,distance)

# All flights by YV planes in month 1 fly the same distance of 229
# there is no variation in the distance travelled by YV carrier flights in month 1 

```

4.  Using `tidyr` and `dplyr` where appropriate, construct a dataframe where the carriers are along the columns, and the rows are the average departure delay (`dep_delay`) flown by each carrier (`carrier`) in each month.\\

```{r}


av_dep_delay <-  flights |> group_by(carrier,month) |> summarise(average.dep.delay = mean(dep_delay, na.rm=TRUE)) |> pivot_wider(names_from = carrier, values_from = average.dep.delay)
av_dep_delay

```

4.  Calculate the proportion of flights that were delayed (`dep_delay` greater than 0) but arrived on or before time (`arr_delay` less than or equal to 0).

    ```{r}

    # extract the flights which meet the criteria 
    flights.relevant <- flights |> filter(dep_delay > 0 & arr_delay <=0) 
    flights.relevant

    # calculate the proportion of flights 
    proportion <- nrow(flights.relevant) / nrow(flights)
    proportion

    ```

5.  Using the `airlines` and `flights` datasets, do the following, showing the output from each step:

-   Identify routes that more than one airline flies

origin, dest : Origin and destination. See airports for additional metadata.

```{r}

# working out for this question 



# origin, dest

# all of the different routes 
routes <- flights |> 
routes


route.groups <- flights |> group_by(origin,dest) |> unique(carrier)
route.groups


# the routes that each carrier takes
carrier.routes <- flights |> group_by(carrier) |> distinct(origin,dest) 
carrier.routes



routes.withcarriers <- flights |> group_by(dest,origin) |> distinct(carrier) |> arrange(dest,origin)

# two ways to count up the number of entires of the dest,origin
routes.withcarriers |> summarise(count=n())
routes.withcarriers |> tally()

number.of.carriers <- routes.withcarriers |> tally()
number.of.carriers

multiple.carriers <- number.of.carriers |> filter(n>1)
multiple.carriers
```

-   Identify routes that more than one airline flies

```{r}

# final answer for this question 

# filter the rows so that it is a data frame with each carrier for each route
routes.withcarriers <- flights |> group_by(origin,dest) |> distinct(carrier, .keep_all=TRUE) 
routes.withcarriers

# select the routes that have multiple carriers 
multiple.carriers <- routes.withcarriers |> tally() |> filter(n>1)
multiple.carriers

```

-   For each such route, calculate the average arrival delay for each airline (exclude NAs). Find the names of these airlines.

```{r}

# extract the flights that are from the routes in the multiple carriers data frame 
# innerjoin shows all of the entries in the first table that have matches in the second table 
mulit.carriers <- flights |> inner_join(multiple.carriers)
mulit.carriers

# calculate the average arrival delay

av_ar_delay <- mulit.carriers |> group_by(origin,dest,carrier) |> summarise("average_arrival_delay"=mean(arr_delay, na.rm=TRUE)) 
av_ar_delay


with_names <- left_join(av_ar_delay,airlines)
with_names

```

-   For each such route, identify the airline with the worst and best average arrival delay.

```{r}
min.max <- with_names|> group_by(origin,dest) |> summarise("min.d" =min(average_arrival_delay), "max.d" = max(average_arrival_delay)) 
min.max




# only want it to macth witht the separate cols ---- make two new cols for best and worst airline names next to the 
left_join(min.max,with_names)

inner_join(min.max,with_names)

middle_join(min.max,with_names)


# doing them separate to get the carriers still in 
min.delay <- with_names|> group_by(origin,dest) |> summarise(min("min" = average_arrival_delay)) |> ungroup()
min.delay 

max.delay <- with_names|> group_by(origin,dest) |> summarise(min("min" = average_arrival_delay), max("max" = average_arrival_delay))
max.delay

# ask how you get it to keep the carrier names? 

```

-   Identify the route with the greatest difference between the best and worst performing airlines

```{r}
ranges <- min.max|> group_by(origin,dest) |> mutate(range.d = max.d-min.d,.before = 1) 
ranges


biggest_range <-ranges |> arrange()
origin_biggest_delay <-biggest_range$origin[1]
dest_biggest_delay <-biggest_range$dest[1]

# alternative could have used summarise ? 
```

The route with the greatest difference between the best and the worst performing airlines was from `origin_biggest_delay` to `dest_biggest_delay`

-   Determine the reason for this difference

```{r}
info <- flights |> filter(origin=="EWR",dest=="ATL") 
info

info.2 <- flights |> filter(origin=="EWR",dest=="ATL") |> distinct(carrier)
info.2
# there are 4 carriers  DL, EV, UA, 9E

relevant.info<- info |> select(!year)
relevant.info

summary(info)

summary(flights)
```

7.  Identify all columns with missing entries, typos and any other inconsistencies in the dataset below (load it just by running the code; created using `dput` command, FYI

```{r}

# load the data 

data.2 <- structure(list(id = c("id_1", "id_2", "id_3", "id_4", "id_5", 
"id_6", "id_7", "id_8", "id_9", "id_10", "id_11", "id_12", "id_13", 
"id_14", "id_15", "id_16", "id_17", "id_18", "id_19", "id_20", 
"id_21", "id_22", "id_23", "id_24", "id_25", "id_26", "id_27", 
"id_28", "id_29", "id_30", "id_31", "id_32", "id_33", "id_34", 
"id_35", "id_36", "id_37", "id_38", "id_39", "id_40", "id_41", 
"id_42", "id_43", "id_44", "id_45", "id_46", "id_47", "id_48", 
"id_49", "id_50"), age = c(50L, 34L, 70L, 33L, 22L, 61L, 69L, 
73L, 62L, 56L, 71L, 33L, 73L, 44L, 45L, 46L, 24L, 70L, 46L, 76L, 
47L, 76L, 28L, 48L, 54L, 27L, 45L, 26L, 61L, 28L, 38L, 55L, 33L, 
36L, 62L, 58L, 72L, 31L, 34L, 51L, 61L, 64L, 26L, 28L, 60L, 29L, 
42L, 46L, 79L, 72L), gender = c("male", "male", "male", "female", 
"female", "male", "female", "male", "male", "female", "female", 
"male", "male", "female", "male", "male", "male", "male", "female", 
"male", "male", "male", "male", "female", "femal", "male", "female", 
"female", "female", "female", "male", "female", "female", "female", 
"male", "male", "female", "male", "female", "female", "male", 
"female", "female", "male", "male", "female", "male", "male", 
"male", "female"), height = c(174.4, 197.7, 174.1, 194.5, NA, 
180.4, 170.5, 157.4, 196.8, 165.1, 153, 197.4, 186, 157.1, 177.5, 
197.7, 179.3, 170.2, 182.4, NA, 165.4, 161, 168.5, 199.2, 157.7, 
154.6, 157.1, 184.5, 181, 194.6, 183.6, 186.9, 176.1, 183, 191.1, 
189.3, 199, 172, 165.6, 170.5, 150.5, 159.2, 192.1, 161.6, 162, 
153.8, 162.3, 186.6, 192.4, 174.9), weight = c(69.4, 62.3, 55.6, 
69.5, 78.6, 60.8, 72.2, 60.9, 75.1, 67.7, 82.5, 68.7, 67.8, 76.7, 
87, 61.1, 70.6, 63.3, 81.5, 59.2, 93.2, 87.3, 83.4, 80.9, 68.6, 
76.5, 93.7, 79.1, 92, 65.6, 85.4, 63.3, 79.7, 74.1, 63.3, 78.2, 
95.7, 95.1, 63.7, 66.1, 99.3, 81, 96.9, 73.3, 70.3, 83, 57.6, 
78.6, 61.9, 98.1), blood_type = c("O", "A", "O", "O", "B", "AB", 
"O", "O", "O", "AB", "A", "O", "O", "O", "B", "A", "B", "AB", 
"O", "AB", "A", "AB", "O", "B", "A", "A", "B", "AB", "A", "B", 
"B", "A", "O", "O", "O", "B", "O", "A", "A", "B", "A", "O", "AB", 
"A", "A", "O", "O", "B", "A", "O"), disease_status = c("diseased", 
"healthy", "healthy", "healthy", "healthy", "healthy", "diseased", 
"healthy", "diseased", "Healthy", "diseased", "healthy", "diseased", 
"healthy", "diseased", "healthy", "healthy", "healthy", "healthy", 
"healthy", "healthy", "diseased", "healthy", "diseased", "healthy", 
"healthy", "healthy", "healthy", "diseased", "diseased", "healthy", 
"healthy", "healthy", "diseased", "diseased", "diseased", "healthy", 
"diseased", "healthy", "healthy", "healthy", "healthy", "healthy", 
"diseased", "diseased", "diseased", "healthy", "healthy", "diseased", 
"diseased"), cholesterol = c(228, 223, 213, 198, 166, 151, 195, 
199, 189, 196, 221, 156, 185, 230, 234, 174, 185, 236, 235, 180, 
165, 220, 160, 153, 250, 153, 184, 242, 212, 179, 224, 233, 181, 
199, 220, 214, 214, 248, 191, 162, 203, 173, 199, 187, 248, 189, 
173, 212, 164, 247), glucose = c(96, 78, 101, 119, 103, 91, 86, 
NA, 77, 80, 115, 85, 88, 109, NA, 71, 90, 94, 91, 87, 113, 93, 
97, 118, 109, 80, 85, 119, 99, 108, 89, 108, 97, 116, 79, 84, 
75, 81, 119, NA, 106, 109, 75, 82, 84, 75, 76, 120, 119, 77), 
    smoker = c("yes", "yes", "yes", "yes", "no", "yes", "no", 
    "yes", "no", "no", "no", "no", "no", "yes", "no", "yes", 
    "yes", "yes", "yes", "yes", "yes", "yes", "yes", "yes", "no", 
    "no", "yes", "yes", "yes", "no", "no", "yes", "no", "yes", 
    "no", "yes", "no", "yes", "yes", "yes", "no", "no", "yes", 
    "no", "no", "no", "no", "no", "no", "yes"), exercise = c("occasional", 
    "regular", "occasional", "regular", "none", "occasional", 
    "regular", "none", "occasional", "none", "occasional", "none", 
    "none", "regular", "occasional", "none", "regular", "regular", 
    "none", "occasional", "none", "occasional", "occasional", 
    "occasional", "regular", "occasional", "regular", "regular", 
    "regular", "occasional", "occasional", "none", "none", "regular", 
    "occasional", "occasional", "none", "none", "none", "none", 
    "occasional", "regular", "regular", "none", "regular", "occasional", 
    "occasional", "none", "occasional", "regular"), income = c(84820L, 
    81547L, 22588L, 72490L, 74533L, 25338L, 41469L, 57315L, 63629L, 
    88662L, 62615L, 56261L, 58499L, 82232L, 77584L, 77275L, 38468L, 
    54510L, 91326L, 78611L, 31402L, 29586L, 21441L, 58269L, 84173L, 
    88295L, 37940L, 43750L, 69750L, 92356L, 82518L, 91455L, 68866L, 
    51178L, 68275L, 27689L, 35418L, 81318L, 62405L, 86851L, 25654L, 
    47553L, 74474L, 51409L, 22607L, 55360L, 96351L, 21516L, 41927L, 
    55810L), education = c("master", "bachelor", "PhD", "master", 
    "bachelor", "highschool", "PhD", "highschool", "PhD", "PhD", 
    "bachelor", "highschool", "master", "bachelor", "PhD", "PhD", 
    "PhD", "bachelor", "master", "highschool", "PhD", "highschool", 
    "bachelor", "master", "highschool", "highschool", "master", 
    "master", "bachelor", "PhD", "highschool", "PhD", "master", 
    "master", "master", "PhD", "highschool", "master", "master", 
    "highschool", "bachelor", "highschool", "bachelor", "PhD", 
    "bachelor", "highschool", "master", "highschool", "bachelor", 
    "bachelor"), region = c("North", "South", "North", "West", 
    "North", "West", "South", "South", "West", "South", "West", 
    "South", "West", "East", "North", "West", "North", "North", 
    "West", "North", "East", "West", "South", "North", "North", 
    "East", "East", "North", "North", "West", "South", "West", 
    "West", "East", "West", "North", "West", "North", "East", 
    "North", "West", "South", "South", "East", "North", "West", 
    "West", "East", "North", "East"), marital_status = c("divorced", 
    "single", "divorced", "divorced", "divorced", "divorced", 
    "divorced", "married", "divorced", "married", "divorced", 
    "widowed", "married", "single", "widowed", "widowed", "single", 
    "divorced", "widowed", "widowed", "single", "married", "single", 
    "married", "widowed", "married", "single", "single", "widowed", 
    "married", "widowed", "divorced", "single", "married", "single", 
    "widowed", "widowed", "married", "widowed", "divorced", "married", 
    "married", "divorced", "single", "married", "widowed", "divorced", 
    "divorced", "single", "divorced")), row.names = c(NA, -50L
), class = c("tbl_df", "tbl", "data.frame"))
```

```{r}

# Identify all columns with missing entries, typos and any other inconsistencies in the dataset

library(skimr)
skim(data.2)


# install the UtilsDataRSV package, use the following code:

if (!requireNamespace("remotes", quietly = TRUE)) {
  install.packages("remotes")
}
remotes::install_github("SATVILab/UtilsDataRSV")
library(UtilsDataRSV)
view_cols(data.2)


#1. gender 
# femal is a type for female 

#2. height 
# there are NA values in height

#3. healthy 
# Healthy" and  "healthy should just be one thing 
# would make Healthy into healthy since everything else is lower case 

# 4. glucose 
# has NA values 
```
